### 2020-01-13 manuscript/manuscript.rmd
update manuscipt

## Abstract
Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages. However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life. Here, we systematically explore the mechanisms (evolutionary models, gene sampling and outgroup choice) that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies. We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. More Importantly, We find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model with inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (especially CAT models). Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of current understanding of the first diverging animal branch and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns. However, a wide consensus of the position of animal has not been reached despite multiple efforts.

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither. As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this effort provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes to explore the root of metazoan phylogeny (Table 1). Although the number of sampled genes might be different across studies, it is reasonable to expect that a conserved core set of orthology genes (OGs) should be present across most animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]). Surprisingly, gene composition and matrix overlap has been rarely assessed across studies and most studies relied on their own bioinformatic pipeline for determination of OGs (Supplementary text) for animal phylogeny.

@@

Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of the root position of animal phylogeny (**Fig XXOverview.C**). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the deep animal phylogeny [@philippe2011resolving]. Feuda *et al*. (2017) [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in data matrices. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. By reanalyzing Chang et al. 2015 [@chang2015genomic] and Whelan et al. 2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved]. However, how the recoding methods impact phylogenetic inference has not been well explored. Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].

@@

```{r overview data matrices}
table1 = select(datasets, -c(paths))
	kable(table1, caption= "Table 1. Overview of data matrices used in this study")

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position that needs to be explored. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ). Importantly, several earlier studies only include less than five ctenophore species. The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other data matrices (Table1, Fi. XXTaxon_composition).

@@

Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogeny. It has been suggested that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination]. However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring]. Thus, it is important to evaluate the gene composition and matrix overlap to explore level of congruence of gene sampling across different matrices. It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones (*e.g.*, Dunn2008 and Hejnol2009).

We first used the Metazoa BUSCO dataset to evaluate gene composition from all data matrices and the result is unexpected. The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco]. Moreover, the BUSCO dataset also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics] ). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, to our surprise, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig. XXBUSCO_annotations).

Previous reanalyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated. Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap). Thus, these results suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.

@@

figure <- ggarrange(p1, p2,
                    labels = c("A", "B"),
                    ncol = 1, nrow = 2, widths=c(1,1), heights=c(1,1),common.legend = TRUE, legend = "right")

@@

**Fig XXGene_composition.** Each of the primary matrices considered here, color coded by the types of genes sampled (XX Ribosomal proteins, etc). Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. support_published_analyses and Table perviouly_published_analyses). Among these `r length(unique(analyses_published$manuscript))` studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (TablexxPAPERS). Moreover, three of them are studies of reanalysis without generating any new data matrices. We found that all past phylogenetic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT (Poisson+CAT+G or GTR+CAT+G) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of the root of animal phylogeny) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT+G and recoding methods with GTR+CAT+G models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig. support_published_analyses).

```{r overview of studies}
table_Papers = read_tsv("../data_processed/tables/previously_published_manuscripts.tsv")
	kable(table_Papers, caption= "Table XX. Overview of previous phylogenomic studies related to the root position of animal phylogeny")

@@

**Fig support_published_analyses.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table perviouly_published_analyses.tsv).

@@

One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we reanalyze the primary matrices from each study under consistent conditions with IQtree under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework. Importantly, we also include a CAT-like, site-heterogeneous C10 - C60 (C) models that are implemented in both ML framework in our analyses and it is interesting to compare results between C model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets). We also progressively trimmed each data matrix with different level of outgroup choice if applicable. We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except only LG+C60+F+G models are used large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.

With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister in every models, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.

@@

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. For computational efficiency, we first only applied Poisson+CAT models in every matrix with Choanoflagellata as outgroup. It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix. Consistent with previous study, we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Nosenko2013 and Whelan2015. Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017. Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model. We found strong support of Porifera-sister in holozoa matrices and Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups. Thus, our results show that both model choice and outgroup sampling play a major role in incongruent results of the root position of animal phylogeny in several key data matrices.

@@

The distribution of phylogenetic signal over the root position of animal phylogeny has never been quantified with site-heterogeneous models. To further explore the effect of model selection and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with different outgroup choice and  models in both ML and BI framework (Fig XXPhylogenetic signal). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

In contrast, the distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in two matrices from Ryan_2013 that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. phylogenetic_signal B). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of root position of animal phylogeny. In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT model and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal. We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models. The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of genes that are supported in each hypothesis across different analyses in Whelan2017 matrices. Consistent with the summary of \Delta$GLS analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of the root of animal phylogeny is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in these key data matrices. Thus, it is clear that recovering Porifera-sister requires more constrains, including more stringent model selection, outgroup choice, and even with gene sampling.

>>>>

## Abstract
Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages. However, a considerable debate still remains regarding to the animal origin, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life. Here, we systematically explore the mechanisms that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies. We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. More Importantly, We find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model with inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (especially CAT models). Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of current understanding of the first diverging animal branch and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns. However, a wide consensus of animl origin has not been reached despite multiple efforts.

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published and extensive reanalyzed, some supporting Ctenophora-sister, some Porifera-sister, and some neither (Table 1). As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this effort provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes to explore the root of animal phylogeny (Table 1). Although the number of sampled genes might be different across studies, it is reasonable to expect that a core set of orthology genes (OGs) should be present and conserved across most animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]). Surprisingly, gene composition and matrix overlap has been rarely assessed across studies and most studies relied on their own bioinformatic pipeline for determination of OGs (for details see Supplementary text section ).

@@

Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of animal evolution (**Fig XXOverview.C**). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the deep animal phylogeny [@philippe2011resolving]. Variation in evolutionary rates across taxa can lead to compositional heterogeneity that may lead to long branch attraction of sponges or ctenophores [@King:2017ie]. Feuda *et al*. [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in matrices. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. By reanalyzing Chang et al. 2015 [@chang2015genomic] and Whelan et al. 2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved]. However, how the recoding methods impact phylogenetic inference has not been well explored. Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six]. Thus, it is important to reevalute the performance of the impact of data-recoding compared to non-recoding methods in animal phylogeny.

@@

```{r overview data matrices}
table1 = read_tsv("../data_processed/tables/study_summary.tsv")
	kable(table1, caption= "Table 1. Overview of data matrices used in this study")

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position that needs to be explored. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]). Importantly, several earlier studies only include less than five ctenophore species (Fig XXTaxon_composition). The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices (Table1, Fi. XXTaxon_composition).

@@

Gene sampling also plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogeny. It has been suggested that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination]. However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring]. Thus, it is important to evaluate the gene composition and matrix overlap to explore level of congruence of gene sampling across different matrices.

We first used the Metazoa BUSCO genes [@waterhouse2017busco0] to evaluate gene composition from all data matrices and the result is unexpected. The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco]. Moreover, the BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig. XXBUSCO_annotations).

Previous analyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated. Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap). Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices. Moreover, the little overlap among studies may potentially lead to incongruent phylogenetic results and difficult to directly compare results across studies.

@@

figure <- ggarrange(p1, p2,
                    labels = c("A", "B"),
                    ncol = 1, nrow = 2, widths=c(1,1), heights=c(1,1),common.legend = TRUE, legend = "right")
figure

@@

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. XXsupport_published_analyses and Table XXperviouly_published_analyses). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (Table 1). Moreover, three of them are reanalyzing studies without generating any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT+G or GTR+CAT+G) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT+G and recoding methods with GTR+CAT+G models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig. support_published_analyses). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups, whereas Ctenophora-sister is supported in all other cases.

@@

**Fig XXsupport_published_analyses.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table XXperviouly_published_analyses).

@@

One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating]. Importantly, we also include a CAT-like, site-heterogeneous C10 - C60 (C) models [@si2008empirical] that are implemented in both ML and Bayesian inference (BI) framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig XXOverview). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60+F+G models are used in 3 large matrices). We found in all cases, WAG/LG+C60+G models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR+C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.

@@

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. For computational efficiency, we first only applied Poisson+CAT+G (Poisson+CAT) models in every matrix with Choanoflagellata as outgroups. It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices. Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013 and Whelan2015. Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large]. The Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and site-homogeneous models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model. We found a weak support of Porifera-sister in Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups and CAT models. Thus, our reanalyzing results are largely consistent with previous analyses, both suggesting that the support of Porifera-sister is increasing with more complex models and closely related outgroups. More importantly, our results indicated that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several key data matrices.

@@

The distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. To further explore the effect of model selection and outgroup choices in animal origin, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with different outgroup choice and models in both ML and BI framework (Fig XXPhylogenetic signal). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS\) or site (\Delta$SLS\) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically comopared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. XXphylogenetic_signal B). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny. In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Fig. XXphylogenetic_signalA, C). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models. The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig XXPairwise_heatmap_animal_root). Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of animal origin is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in Whelan2017_full matrices.
